---
title: "SISCER Survival Course R Practicum Day 1"
author: "Pamela Shaw"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, warning = F, message = F}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(ggsurvfit)
library(km.ci)
library(survminer) #ggsurvplot
library(gtsummary) #survfit table
library(lmtest) # likelihood ratio test

```

## Dataset: `veteran` from the **survival** package

```{r load-data}
data(cancer, package = "survival")

# convert days to months
veteran$months <- veteran$time / (365.25 / 12)  
head(veteran)
```

## Exercises

1. **Kaplan-Meier Curves**: Plot Kaplan-Meier survival curves by cell type. 
    * Make one version with **pointwise confidence intervals**.  
    * Make another with **simultaneous confidence bands**. 

2. **Log-Rank Test**: Compute and interpret the log-rank statistic comparing survival by cell type.

3. **Assumptions**: What assumption is required for the log-rank test to be most efficient?

4. **Check Proportional Hazards**: Check the proportional hazards assumption.  
    * Graphically using log(-log(S(t))).  
    * Using a formal test via `cox.zph`.  
    * Conclude whether the log-rank test is reasonable in this case.

5. **Peto-Peto Test**: Compute the Peto-Peto test and compare results to part 2.

6. **Improved Visualization**: Revisit your Kaplan-Meier curves and improve the plot aesthetics.  
    * Add confidence intervals.  
    * Include a risk table.  
    * Modify color, font size, or confidence interval format (shaded vs lines).

7. **Restricted Mean Survival**: Compare the restricted mean survival times between cell types, restricted to 1 year.

8. **Cox Proportional Hazards Model**: Fit a multivariable Cox model adjusting for cell type, treatment, prior therapy, and Karnofsky score.  
    i. Do the HRs for cell type summarize the stratified K-M curves well?  
    ii. What is the HR for adeno vs squamous cell type?  
    iii. How important is cell type in the model? Compare this to the unadjusted log-rank result.  
    iv. What is the HR associated with a 10-point increase in Karnofsky score?  
    v. Evaluate the PH assumption using `cox.zph`.  
    vi. Refit the model stratifying on cell type. Are any PH violations still present?
    
## Answers 

### 1. Kaplan-Meier Curves

#### Pointwise confidence bands 

```{r Q1a}
vfit.surv <- survfit(Surv(months, status) ~ celltype, data = veteran, conf.type = "log-log")
survfit2(Surv(months, status) ~ celltype, data = veteran) %>%
  ggsurvfit() +
  labs(
    x = "Months",
    y = "Overall survival probability"
  ) +
  add_confidence_interval() 
```

#### Simultaneous confidence bands

```{r Q1b}
# Subset to one cell type, e.g., "squamous"
# Can repeat for each cell type 
squamous_data <- subset(veteran, celltype == "squamous")

# Create surv object
surv_simul <- survfit(Surv(months, status) ~ 1, data = squamous_data)

# Compute simultaneous confidence band
fit.simul <- km.ci(surv_simul, method = "logep")

# Plot
plot(fit.simul, xlim = c(0, 12), xlab = "Months", ylab = "Survival Probability")
title("Simultaneous logEP 95% CI: Squamous")


```

### 2. Log-Rank Test

The log-rank test for equality of survival curves across cell types yielded a 
p-value < 0.001, indicating strong evidence of a significant difference in survival distributions among the cell types.

```{r Q2}
survdiff(Surv(months, status) ~ celltype, data = veteran)
```

### 3. Assumptions

The log-rank test assumes proportional hazards, meaning the hazard ratios between groups are constant over time.

### 4. Check Proportional Hazards

The complementary log-log curves are not parallel, suggesting a violation to the 
proportional hazards assumption. In addition, the score test from cox.zph 
(p = 0.03) rejects the proportional-hazards assumption.

```{r Q4}
# Graphical check
plot(vfit.surv, fun = "cloglog", col = 1:4, xlab = "log(t)", ylab = "log(-log(S(t)))", main = "Examining log hazard")
legend("bottomright", legend = levels(veteran$celltype), col = 1:4, lty = 1, bty = "n")

# Formal test
vfit.cox.unadj <- coxph(Surv(months, status) ~ celltype, data = veteran)
vfit.ph <- cox.zph(vfit.cox.unadj,terms=F)
print(vfit.ph)
plot(vfit.ph, hr = TRUE)
### Note we added "terms=F" so that we would get a plot for each coefficient. Without that in the cox.zph statement there would only one plot for cell type, and an overall pvalue for that variable.
```

### 5. Peto-Peto Test

Like the log-rank test, the Peto-Peto test for equality of survival curves 
across cell types yielded a p-value < 0.001, indicating strong evidence of a significant difference in survival distributions among the cell types.

```{r Q5}
survdiff(Surv(months, status) ~ celltype, data = veteran, rho = 1)
```

### 6. Improved Visualization

```{r Q6}
survfit2(Surv(months, status) ~ celltype, data = veteran) %>%
  ggsurvfit() +
  add_confidence_interval(type = "ribbon") +
  add_risktable() +
  labs(title = "Survival by Cell Type", x = "Months", y = "Survival Probability")
```

### 7. Restricted Mean Survival

```{r Q7}
print(vfit.surv, rmean = 12)

### IF we wanted to compare two groups, we could form our own t-test by hand, following code we did in Lecture t=diff/se. (I dont know of an easier way in R, but I'm sure there must be). 

# for differnece or the RMST (rmean) between squamous and small cell have
diff=5.61-2.55
## since independent groups, add variance to compute SE for difference
se.diff<-sqrt(.797^2 + .459^2)
Zstat=diff/se.diff
pvalue <- 2 * (pnorm(abs(Zstat), lower.tail = FALSE))
pvalue
### Note this would be more precise as t-statistic but would need to calculate Saitterthwhaite's formula to calculate df by hand.
## https://www.statisticshowto.com/satterthwaite-formula/
##Or perhaps instead find an R package to do this test of difference for RMST properly for us!

```

### 8. Cox Proportional Hazards Model

#### Part (i)

Yes, the HRs for cell type summarize the stratified K-M curves well; for example, 
the adeno group has the highest hazard ratio, and we can visually see from the 
K-M plots that it has the fastest drop in survival. 

```{r Q8i}
# Multivariable Cox model
vfit.cox.mv <- coxph(Surv(months, status) ~ celltype + trt + prior + karno, data = veteran)
summary(vfit.cox.mv)
```

#### Part (ii)

The hazard ratio for adeno vs squamous cell type is 3.24. 

#### Part (iii)

```{r Q8iii}
# Likelihood ratio test (LRT) for cell type
vfit.cox.mv0 <- coxph(Surv(months, status) ~ trt + prior + karno, data = veteran)
lrtest(vfit.cox.mv0, vfit.cox.mv)

# unadjusted log rank result
survdiff(Surv(months, status) ~ celltype, data = veteran)

```
The p-value from LRT from multivariate model and logrank test from univariable model were very similar, both <0.001. The unadjusted test was a little smaller, which isnt that suprising as some of its significance may have been due to correlation with other variables left out of the model 

#### Part (iv)

The HR associated with a 10-point increase in Karnofsky score is $exp(-0.0315 * 10) = 0.73$.

You could get confidence interval for this quantity in this way (note I am sure there is a function in R to find the confidence interval for a function of a regression coefficient. Perhpas one of you could find it. But for a monotonic transformation, it's easy to do confidence interval for 10*beta on log-scale and then exponentiate.

```{r CI_HR_10ptincrease}
beta.karno<-summary(vfit.cox.mv)$coef[6,1]
se.beta.karno<-summary(vfit.cox.mv)$coef[6,3]
LL<-10*beta.karno - qnorm(.975)*10*se.beta.karno
UL<-10*beta.karno + qnorm(.975)*10*se.beta.karno

round(c(exp(LL),exp(UL)),3)
```

#### Part (v)

```{r Q8v}
# PH assumption
vfit.mv.ph <- cox.zph(vfit.cox.mv,terms=F)
print(vfit.mv.ph)
```

#### Part (vi)

```{r}
# Stratified model
vfit.cox.strat <- coxph(Surv(months, status) ~ trt + prior + karno + strata(celltype), data = veteran)
summary(vfit.cox.strat)

vfit.strat.ph <- cox.zph(vfit.cox.strat)
print(vfit.strat.ph)
```
From these results, we see some residual violations of the PH assumption, but things got a bit better.